Handling Recursion in Generic Programming Using Closed Type Families

### 2018-04-26 lncs-paper.tex
Polish

The following exploits the idea of \emph{pattern matching}. As before, we consider two cases. In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \texttt{Show}, and invoke its \texttt{show} function. Whereas in the case of recursion point, we surround it with parentheses and apply our generic function \texttt{gshow}. Thus, the type family for equality model a form of pattern matching on the types again:
\texttt{
\vs
\indent\textbf{class} CaseShow (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s caseShow' \ann b \ra String
\vs
\indent\textbf{instance} Show\s\s b \Ra CaseShow 'False a b \textbf{where}\\
\indent\s\s caseShow'\s\s\s = show\\
\indent\textbf{instance} GShow a \Ra CaseShow 'True\s\s a a \textbf{where}\\
\indent\s\s caseShow' t = "(" ++ gshow t ++ ")"
\vs
}
We provide a synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, as before with \texttt{ProofEq}; likewise the synonym for the matching function:
\texttt{
\vs
\indent caseShow \ann \fa a b. CaseRecShow a b \Ra b \ra String\\
\indent caseShow t = caseShow' @(Equal a b) @a t
\vspace{0.2cm}
}

@@

Using the defined operations, we can implement differentiation of a product of types. The definition of differentiation shown in Figure~\ref{fig:diff-prod} resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \texttt{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$. When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \texttt{Hole}. Reflecting this type in the context is helpful when we traverse the context representation to plug the hole. We use the empty type \texttt{End} to distinguish the case when the hole is found at the end of the list, in order not to add that into the result twice. The sum \texttt{'[ '[]]} represents a unit type that is exactly $1$. We also use type-level \texttt{If} that returns its second argument for \texttt{'True}, and the third one otherwise. We do not give its definition here, as it is straightforward.
\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt data Hole = Hole\\
\tt data End
\vs
\tt type family DiffProd (fam \ann [*]) (xs \ann [*]) \ann [[*]] where\\
\tt\s\s DiffProd fam '[]\ind\s\s\s = '[]\\
\tt\s\s DiffProd fam '[x]\ind\s\s = If (InFam x fam) '[ '[Hole]] '[]\\
\tt\s\s DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]\ind\s '[]\\
\tt\s\s DiffProd fam (x ': xs)\\
\tt\ind = Hole .* xs .** DiffProd fam '[End, x]  .++\\
\tt\ind\ind\ind\s\s x\s\s .*\s\s DiffProd fam xs
\end{tabular}
\caption{Differentiation of a product of types.}
\label{fig:diff-prod}
\end{figure}

@@

We finally define \texttt{Context} as a newtype that wraps the result of the computation, as it allows GHC to perform type inference for the context type, where it is possible, to avoid extra type applications for defining the zipper interface functions.
\texttt{
\vs
\indent type CtxCode  fam a = ToContext 'F fam (Code a)
\vs
\indent newtype Context fam a = Ctx \{ctx \ann SOP I (CtxCode fam a)\}
\vs
}
The \texttt{CtxCode} type is the computed type of the context for the given code of a datatype. We will use this type synonym further when defining constraints for functions providing the generic zipper interface.

>>>>

The following exploits the idea of \emph{pattern matching}. As before, we consider two cases. In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \texttt{Show}, and invoke its \texttt{show} function. Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \texttt{gshow}. Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
\texttt{
\vs
\indent\textbf{class} CaseShow (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s caseShow' \ann b \ra String
\vs
\indent\textbf{instance} Show\s\s b \Ra CaseShow 'False a b \textbf{where}\\
\indent\s\s caseShow'\s\s\s = show\\
\indent\textbf{instance} GShow a \Ra CaseShow 'True\s\s a a \textbf{where}\\
\indent\s\s caseShow' t = "(" ++ gshow t ++ ")"
\vs
}
We provide a synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, as before with \texttt{ProofEq}; likewise the synonym for the matching function:
\texttt{
\vs
\indent caseShow \ann \fa a b. CaseRecShow a b \Ra b \ra String\\
\indent caseShow t = caseShow' @(Equal a b) @a t
\vspace{0.2cm}
}

@@

Using the defined operations, we can implement differentiation of a product of types. The definition of differentiation shown in Figure~\ref{fig:diff-prod} resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \texttt{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$. When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \texttt{Hole}. Reflecting this type in the context is helpful when we traverse the context representation to plug the hole. We use the empty type \texttt{End} to distinguish the case when the hole is found at the end of the list, in order not to add that into the result twice. The sum \texttt{'[ '[]]} represents a unit type that is exactly $1$. We also use type-level \texttt{If} that returns its second argument for \texttt{'True}, and the third one otherwise. We do not give its definition here, as it is straightforward.
\begin{figure}[!b]
\centering
\normalsize
\begin{tabular}{l}
\tt data Hole = Hole\\
\tt data End
\vs
\tt type family DiffProd (fam \ann [*]) (xs \ann [*]) \ann [[*]] where\\
\tt\s\s DiffProd fam '[]\ind\s\s\s = '[]\\
\tt\s\s DiffProd fam '[x]\ind\s\s = If (InFam x fam) '[ '[Hole]] '[]\\
\tt\s\s DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]\ind\s '[]\\
\tt\s\s DiffProd fam (x ': xs)\\
\tt\ind = Hole .* xs .** DiffProd fam '[End, x]  .++\\
\tt\ind\ind\ind\s\s x\s\s .*\s\s DiffProd fam xs
\end{tabular}
\caption{Differentiation of a product of types.}
\label{fig:diff-prod}
\end{figure}

@@

We finally define \texttt{Context} as a newtype wrapping the result of the computation, as it allows GHC to perform type inference for the context type, where it is possible, to avoid extra type applications for defining the zipper operations.
\texttt{
\vs
\indent type CtxCode  fam a = ToContext 'F fam (Code a)
\vs
\indent newtype Context fam a = Ctx \{ctx \ann SOP I (CtxCode fam a)\}
\vs
}
The \texttt{CtxCode} type is the computed type of the context for the given code of a datatype. We will use this type synonym further when defining constraints for functions providing the generic zipper interface.
